---
title: "Drawing ROC Curves in R with ROCR Package"
author: "Yao-Jen Kuo"
date: "Monday, September 21, 2015"
output: 
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
    theme: united
    highlight: tango
    fig_width: 9
    fig_height: 5
---

#Introduction
ROC curves are commonly used in the performance evaluation of models with binary outcomes. This article demonstrates an easy example of building logistic regression with R's built-in dataset "diamonds" and drawing ROC curves to determine which characteristic: carat, cut, or clarity best predicts whether the diamond is expensive or not.
We will use `library(ggplot2)` and `library(ROCR)` these 2 packages in this article.

#Exploring and Preparing Data
Let us first check the structure and statistics summary of "diamonds."
```{r, echo=FALSE}
library(ROCR)
library(ggplot2)
```
```{r, echo=TRUE}
str(diamonds)
summary(diamonds)
```

##Creating a Binary Target: isExpensive
There is no binary variable for our target in the original "diamonds", so let us assign that price over median: $2,400 as "EXPENSIVE."
```{r, echo=TRUE}
plot(density(diamonds$price), main="Distribution of Diamond Price")
polygon(density(diamonds$price), col="pink")
abline(v=2400, lwd=3, col="skyblue")
diamonds$isExpensive <- diamonds$price > 2400
```

#Training a Model on The Data
Since the number of samples are abundant, as usual, we adopt 70% for training and 30% for test.
```{r}
set.seed(1)
isTest <- runif(nrow(diamonds)) > 0.70
train <- diamonds[isTest==FALSE,]
test <- diamonds[isTest==TRUE,]
```

##Model 1: cut as the only predictor
```{r, echo=TRUE}
fit1 <- glm(isExpensive ~cut, data=train, family=binomial())
```
```{r, echo=FALSE}
prob1 <- predict(fit1, newdata=test, type="response")
pred1 <- prediction(prob1, test$isExpensive)
perf1 <- performance(pred1, measure = "tpr", x.measure = "fpr")
auc1 <- performance(pred1, measure = "auc")
auc1 <- auc1@y.values[[1]]
```
```{r, echo=TRUE}
auc1
```

##Model 2: clarity as the only predictor
```{r, echo=TRUE}
fit2 <- glm(isExpensive ~clarity, data=train, family=binomial())
```
```{r, echo=FALSE}
prob2 <- predict(fit2, newdata=test, type="response")
pred2 <- prediction(prob2, test$isExpensive)
perf2 <- performance(pred2, measure = "tpr", x.measure = "fpr")
auc2 <- performance(pred2, measure = "auc")
auc2 <- auc2@y.values[[1]]
```
```{r, echo=TRUE}
auc2
```

##Model 3: carat as the only predictor
```{r, echo=TRUE}
fit3 <- glm(isExpensive ~carat, data=train, family=binomial())
```
```{r, echo=FALSE}
prob3 <- predict(fit3, newdata=test, type="response")
pred3 <- prediction(prob3, test$isExpensive)
perf3 <- performance(pred3, measure = "tpr", x.measure = "fpr")
auc3 <- performance(pred3, measure = "auc")
auc3 <- auc3@y.values[[1]]
```
```{r, echo=TRUE}
auc3
```

#Evaluating Model Performance
Drawing ROC curves.
```{r, echo=TRUE}
plot(perf1, colorize = TRUE, lwd=3, main="ROC Curves")
plot(perf2, add = TRUE, colorize = TRUE, lwd=3)
plot(perf3, add = TRUE, colorize = TRUE, lwd=3)
abline(a=0, b=1, lty=2, lwd=3, col="black")
```

#Conclusion
* Size matters. Using $carat as the only predictor can achieve a AUC of 99%.
* We can easily build logistic regression models using `fit <- glm()` with `family=binomial()` as assigned parameter.
* We also demostrated how to generate training/test datasets from the original dataset in R.
* Most importantly, we demonstrated how to plot beautiful ROC curves.